Presheaf Model for Dependent Types

### 2017-09-30 papers/talk-11/paper-11.tex
Fix example for A -> A
Correct wording/typos

Suppose you want to write a function that returns the head of a given list of
natural numbers, using the classical definition of the list type:
\[
  List = Nil : Unit \to List~|~Cons : \NN \times List \to List
\]
Writing the case for $Cons$ is easy --- just take the first element of the
pair.  But what about the case of an empty list? This case is tricky because it
is not clear what should the result be. Obviously, such function (if we do not
allow definition of partial functions which is the case in many programming
languages) cannot be well defined on the empty list without taking arbitrary
decisions. But maybe such problems can be prevent from happening at all?

A solution to this issue is an instance for the application of dependent types.
Dependent types are such types that include simple types, like products and
functions, that we have seen in simply-typed lambda calculus (STLC), but also
add a possibility to define types using values. With dependent types we would
be able to define a type for lists that includes the size of the list:
\[
  List(\NN) = Nil : Unit \to List(0)~|~Cons : \NN \times List(n) \to List(Suc(n))
\]
\begin{example}
  The function to construct a list of length $n$, having only $0$ as elements
  would then have the type $\Pi n: \NN.List(n)$, indicating that it accepts a
  natural number and returns a value of a type constructed using that number.
\end{example}
\begin{example}
  Our desired head function would have the type
  \[
    Hd: \Pi n: \NN.List(Suc(n)) \to \NN
  \]
  indicating that it cannot be applied to the empty list at all.
\end{example}

@@

To formalize dependent types we will use a notion of typing context (which is
almost the same as typing environment as seen for STLC, with the exception that
it holds dependent types).

\begin{definition}[Context]
  A \emph{context} is a list of bindings that connect variables to types: $x_1: A_1,
  \dots, x_n: A_n$. A context is called well-formed if every variable in it is
  distinct from the other variables and each $A_i$ is a type in the context $x_1
  : A_1, \dots, x_{n-1} : A_{n-1}$, since our types may now include the variables
  from the context.
\end{definition}

Note that we extend the notion of a well-formed context from STLC --- in STLC
we only needed the variables to be distinct, since they had no influence on the
types.

@@

\begin{definition}[Structural rules]
\begin{mathpar}
  \inferrule*[right=C-Emp]{ }{\vdash \diamond~ctxt}
  \and
  \inferrule*[right=C-Ext]{\Gamma \vdash A~type}{\vdash \Gamma, x : A~ctxt}
  \and
  \inferrule*[right=C-Ext-Eq]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\vdash \Gamma, x:A = \Delta, y:B~ctxt}
  \and
  \inferrule*[right=Var]{\vdash \Gamma, x:A,\Delta~ctxt}{\Gamma,x:A,\Delta \vdash x:A}
  \and
  \inferrule*[right=C-Eq-R]{\vdash \Gamma~ctxt}{\vdash \Gamma = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-S]{\vdash \Gamma = \Delta~ctxt}{\vdash \Delta = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-T]{\vdash \Gamma = \Delta~ctxt \\ \vdash \Delta = \Theta~ctxt}{\vdash \Gamma = \Theta~ctxt}
  \and
  \inferrule*[right=Tm-Conv]{\Gamma \vdash a : A \\ \vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\Delta \vdash a : B}
  \and
  \inferrule*[right=Ty-Conv]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A~type}{\Delta \vdash A~type}
  \and
  \inferrule*[right=Weak]{\Gamma, \Delta \vdash \mathcal{J} \\ \Gamma \vdash C~type}{\Gamma, x : C, \Delta \vdash \mathcal{J}}
  \and
  \inferrule*[right=Subst]{\Gamma, x: C, \Delta \vdash \mathcal{J} \qquad \Gamma \vdash c: C}{\Gamma, \Delta[c/x] \vdash \mathcal{J}[c/x]}
\end{mathpar}

where $x$ and $y$ in C-Ext and C-Ext-Eq are assumed to be fresh variables and
$\mathcal{J}$ in Weak and Subst ranges over judgements of the form $a : A,
A~type, a = b: A, A = B~type$. $[c/x]$ denotes the same substitution as seen in
STLC, except that it now also applies to the variables that may be contained in
types.
\end{definition}

@@

Rules C-Eq define that context equality is an equivalence relation, very
similar rules for terms and types, stating that their equalities are also
equivalence relations, are omitted for brevity.

@@

To introduce instances of $\Pi$-type we use the rule $\Pi$-I which looks almost
the same as in the simply-typed case, with the exception that our $\lambda$
here forms a dependent function type, not the simply-typed one.

The rule $\Pi$-I-Eq serves similar purpose to $\Pi$-Eq and ensures the
preservation of the definitional equality. The similar equality preservation
rules henceforth will be omitted for the other rules we introduce, but the
reader should keep their implicit presence in mind while seeing other term and
type formers that we introduce.

The rule $\Pi$-E regulates the application of a dependent function to some value
of the correct type results in the return value of a proper type. Note the rule
only concerns the resulting type here, not the value and that the bound
variable $x$ (shown as $[x : A] B$ in subscript) is replaced in the type, not
the value, which is handled by the next rule.

@@

The formation rule $\Sigma$-F follows the same pattern as the rule for
dependent function, but the introduction is a bit different. Note that we take
such $b$, where the type contains a variable $x : A$ that is substituted by the
value $a$. Together the values $a$ and $b$ form a pair, where the type of the
second element depends on the value of the first. We use the notation $[x : A]
B$ here again to indicate that the variable $x$ is bound in the type $B$.

@@

We know the projection functions for the simply-typed products in STLC. Let us
now define the projection functions for the \emph{dependent} products, which we
call $\mathbf{p}$ for the first projection and $\mathbf{q}$ for the second
projection:
\[
  \mathbf{p}(a) := R^\Sigma_{[z : \Sigma x: A. B]A}([x : A, y: B]x, a) : A
\]
\[
  \mathbf{q}(a) := R^\Sigma_{[z : \Sigma x: A. B]B[z.1/x]}([x : A, y: B]y, a) : B[\mathbf{p}(a)/x]
\]
where $\Gamma \vdash A~type$, $\Gamma, x:A \vdash B~type$, and $\Gamma \vdash a
: \Sigma x: A. B$.

The analogue to dependent sum in set theory is a disjoint union. For every
family of sets $(B_i)_{i \in I}$ we can form the set $\Sigma_{i \in I} B_i :=
\{ (i, b)~|~i \in I \land b \in B_i\}$ whose elements consist of an index $i$
and an element of the corresponding set $B_i$.

@@

A note on notation: we sometimes write $A\sigma$, where $\Gamma \vdash
A\sigma~type$, $\Delta \vdash A~type$, and $\sigma \from \Gamma \to \Delta$ for
some contexts $\Gamma$ and $\Delta$, to denote the application of $\sigma$ on
the right. That is, $A\sigma$ becomes here a type in the domain context of
$\sigma$, where $\sigma(A\sigma) = A$. The same notation is sometimes
analogously used for terms: $a\sigma$ to denote a term $\Gamma \vdash a\sigma :
A\sigma$, where $\Delta \vdash a : A$.

Having this notation, we also want the following equations to hold:
\[
  A1=A \quad (A\sigma)\tau = A(\sigma\tau) \quad a1=a \quad (a\sigma)\tau=a(\sigma\tau)
\]
For readability, we usually omit writing indices to the identity morphism $1$,
using it as the most suitable identity morphism for the surroundings. We also
assume correct types in the equations, i.e. $(A\sigma)\tau = A(\sigma\tau)$
assumes $\sigma \from \Delta \to \Gamma$, $\tau \from \Theta \to \Delta$, and
$\Gamma \vdash A~type$.

@@

There are several ways to model dependent types with category theory, the one
we show is using Categories with Families.

@@

Let us see how the category of presheaves over $\cat{C}$ (we denote it as
$Psh(\cat{C})$) forms a category of families where presheaves are contexts.  If
we set presheaves to be the contexts, then the context morphisms are morphisms
in $[\cat{C}^\op, \Set]$, which are natural transformations between contravariant
functors. These natural transformations satisfy the following naturality
condition:
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]
where $I, J \in \ob(\cat{C}); \Gamma, \Delta \in Psh(\cat{C}); \sigma
\from \Delta \to \Gamma$, and $f \from J \to I$. $\Gamma_f$ and $\Delta_f$
denote the morphisms that we obtain from the morphism $f$ in $\Set$ by the
functoriality of $\Gamma$ and $\Delta$. For instance, if $\rho \in \Gamma(I)$,
then we have a morphism $\rho \mapsto \rho f$ written as $\Gamma_f$ to the
corresponding $\rho f \in \Gamma(J)$, (we write $f$ applied on the right to
connect to the initial morphism in $\cat{C}$) since the contravariant functor
$\Gamma$ preserves morphism mappings.

Recall that every context has a morphism from itself to the empty context.
Hence we set the empty context to be the terminal object in the category of
presheaves, the terminal context, which is a functor that maps everything to
the terminal object in $\Set$, the singleton set. Indeed, we need a unique
natural transformation from every presheaf to the terminal presheaf. This
natural transformation is a family of unique morphisms to the terminal object
in $\Set$.

@@

The second projection $\mathbf{q}$ here is the one that models the variables in
an extended context. The application of it returns the current term for the
variable. Note the similarity to the variable rule in the ccc model of STLC, as
it is also modeled by a second projection there.

@@

\begin{example}
  Let us consider a representation of a term of type $A \to A$ for some $\vdash
  \Gamma~ctxt$, $\Gamma \vdash A~type$ in the presheaf model. For a CwF $(\cat{C}, F), I, J \in ob(\cat{C}), f \from J \to I$ we give the
  element of $\mathit{Ty}(\Gamma)$ corresponding to the type $A \to A$:
  \[
    (A \to A)\rho = \{\pi~|~\pi_f(u) \in A(\rho f), u \in A(\rho f)\} = \{\lambda I f u.u\}
  \]
  Note that the required equalities hold trivially for any $g \from K \to J$, $((\lambda
  I f u.u)_f u) g = (\lambda I f u.u)_{fg} (ug)$. $\Gamma, x : A \vdash x : A$ corresponds to $q$ in the model. Thus $\Gamma
  \vdash \lambda x.x : A -> A$ corresponds to $\lambda q$ being defined as
  follows:
  \[
    ((\lambda q) \rho)_f u = q(\rho f, u) = u
  \]
\end{example}

>>>>

Suppose you want to write a function that returns the head of a given list of
natural numbers, using the classical definition of the list type:
\[
  List = Nil : Unit \to List~|~Cons : \NN \times List \to List
\]
Writing the case for $Cons$ is easy --- just take the first element of the
pair.  But what about the case of an empty list? This case is tricky because it
is not clear what should the result be. Obviously, such function (if we do not
allow definitions of partial functions which is the case in many programming
languages) cannot be well defined on the empty list without taking arbitrary
decisions. But maybe such problems can be prevented from happening at all?

A solution to this issue is an instance for the application of dependent types.
Dependent types are such types that include simple types, like products and
functions, that we have seen in simply-typed lambda calculus (STLC), but also
add a possibility to define types using values. With dependent types we would
be able to define a type for lists that includes the size of the list:
\[
  List(\NN) = Nil : Unit \to List(0)~|~Cons : \NN \times List(n) \to List(Suc(n))
\]
\begin{example}
  The function to construct a list of a length $n$, having only $0$ as elements
  would then have the type $\Pi n: \NN.List(n)$, indicating that it accepts a
  natural number and returns a value of a type constructed using that number.
\end{example}
\begin{example}
  Our desired head function would have the type
  \[
    Hd: \Pi n: \NN.List(Suc(n)) \to \NN
  \]
  indicating that it cannot be applied to the empty list at all.
\end{example}

@@

To formalize dependent types we will use a notion of a typing context (which is
almost the same as the typing environment as seen for STLC, with the exception
that it holds dependent types).

\begin{definition}[Context]
  A \emph{context} is a list of bindings that connect variables to types: $x_1: A_1,
  \dots, x_n: A_n$. A context is called well-formed if every variable in it is
  distinct from the other variables and each $A_i$ is a type in the context $x_1
  : A_1, \dots, x_{n-1} : A_{n-1}$.
\end{definition}

Note that we extend the notion of a well-formed context from STLC --- in STLC
we only needed the variables to be distinct, since they had no influence on the
types. We require the additional condition because our types may now include
the variables from the context.

@@

\begin{definition}[Structural rules]
\begin{mathpar}
  \inferrule*[right=C-Emp]{ }{\vdash \diamond~ctxt}
  \and
  \inferrule*[right=C-Ext]{\Gamma \vdash A~type}{\vdash \Gamma, x : A~ctxt}
  \and
  \inferrule*[right=C-Ext-Eq]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\vdash \Gamma, x:A = \Delta, y:B~ctxt}
  \and
  \inferrule*[right=Var]{\vdash \Gamma, x:A,\Delta~ctxt}{\Gamma,x:A,\Delta \vdash x:A}
  \and
  \inferrule*[right=C-Eq-R]{\vdash \Gamma~ctxt}{\vdash \Gamma = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-S]{\vdash \Gamma = \Delta~ctxt}{\vdash \Delta = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-T]{\vdash \Gamma = \Delta~ctxt \\ \vdash \Delta = \Theta~ctxt}{\vdash \Gamma = \Theta~ctxt}
  \and
  \inferrule*[right=Tm-Conv]{\Gamma \vdash a : A \\ \vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\Delta \vdash a : B}
  \and
  \inferrule*[right=Ty-Conv]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A~type}{\Delta \vdash A~type}
  \and
  \inferrule*[right=Weak]{\Gamma, \Delta \vdash \mathcal{J} \\ \Gamma \vdash C~type}{\Gamma, x : C, \Delta \vdash \mathcal{J}}
  \and
  \inferrule*[right=Subst]{\Gamma, x: C, \Delta \vdash \mathcal{J} \qquad \Gamma \vdash c: C}{\Gamma, \Delta[c/x] \vdash \mathcal{J}[c/x]}
\end{mathpar}
where $x$ and $y$ in C-Ext and C-Ext-Eq are assumed to be fresh variables and
$\mathcal{J}$ in Weak and Subst ranges over the judgements of the form $a : A,
A~type, a = b: A, A = B~type$. $[c/x]$ denotes the same substitution as seen in
STLC, except that it now also applies to the variables that may be contained in
the types.
\end{definition}

@@

Rules C-Eq define that context equality is an equivalence relation. Very
similar rules for terms and types, stating that their equalities are also
equivalence relations, are omitted for brevity. One should keep in mind however
that we implicitly include them too.

@@

To introduce instances of $\Pi$-type we use the rule $\Pi$-I which looks almost
the same as in the simply-typed case, with the exception that our $\lambda$
here forms a dependent function type, not a simply-typed one.

The rule $\Pi$-I-Eq serves similar purpose to $\Pi$-Eq and ensures the
preservation of the definitional equality. The similar equality preservation
rules henceforth will be omitted for the other rules we introduce, but the
reader should keep their implicit presence in mind while seeing other term and
type formers.

The rule $\Pi$-E regulates that the application of a dependent function to some
value of a correct type results in the return value of a proper type. Note that
the rule only concerns the resulting type here --- not the value --- and that
the bound variable $x$ (shown as $[x : A] B$ in subscript) is replaced in the
type, not the value, which is handled by the next rule.

@@

The formation rule $\Sigma$-F follows the same pattern as the rule for
dependent functions, but the introduction is a bit different. Note that we take
such $b$, where the type contains a variable $x : A$ that is substituted by the
value $a$. Together the values $a$ and $b$ form a pair, where the type of the
second element depends on the value of the first. We use the notation $[x : A]
B$ here again to indicate that the variable $x$ is bound in the type $B$.

@@

We know the projection functions for the simply-typed pairs in STLC. Let us now
define the projection functions for the \emph{dependent} pairs, which we call
$\mathbf{p}$ for the first projection and $\mathbf{q}$ for the second
projection:
\[
  \mathbf{p}(a) := R^\Sigma_{[z : \Sigma x: A. B]A}([x : A, y: B]x, a) : A
\]
\[
  \mathbf{q}(a) := R^\Sigma_{[z : \Sigma x: A. B]B[z.1/x]}([x : A, y: B]y, a) : B[\mathbf{p}(a)/x]
\]
where $\Gamma \vdash A~type$, $\Gamma, x:A \vdash B~type$, and $\Gamma \vdash a
: \Sigma x: A. B$.

The analogue to dependent sum in set theory is disjoint union. For every family
of sets $(B_i)_{i \in I}$ we can form the set $\Sigma_{i \in I} B_i := \{ (i,
b)~|~i \in I \land b \in B_i\}$ whose elements consist of an index $i$ and an
element of the corresponding set $B_i$.

@@

We sometimes write $A\sigma$, where $\Gamma \vdash A\sigma~type$, $\Delta
\vdash A~type$, and $\sigma \from \Gamma \to \Delta$ for some contexts $\Gamma$
and $\Delta$, to denote the application of $\sigma$ on the right. That is,
$A\sigma$ becomes here a type in the domain context of $\sigma$, where
$\sigma(A\sigma) = A$. The same notation is sometimes analogously used for
terms: $a\sigma$ denotes a term $\Gamma \vdash a\sigma : A\sigma$, where
$\Delta \vdash a : A$.

Having this notation, we also want the following equations to hold:
\[
  A1=A \quad (A\sigma)\tau = A(\sigma\tau) \quad a1=a \quad (a\sigma)\tau=a(\sigma\tau)
\]
For readability, we usually omit writing indices to the identity morphism $1$,
using it as the most suitable identity morphism for the surroundings. We also
assume the correct types in the equations, i.e. $(A\sigma)\tau = A(\sigma\tau)$
assumes $\sigma \from \Delta \to \Gamma$, $\tau \from \Theta \to \Delta$, and
$\Gamma \vdash A~type$.

@@

There are several ways to model dependent types with category theory, the one
we show relies on Categories with Families.

@@

Let us see how the category of presheaves over $\cat{C}$ (we denote it as
$Psh(\cat{C})$) forms a category of families where presheaves are contexts. If
we set presheaves to be the contexts, then the context morphisms are morphisms
in $[\cat{C}^\op, \Set]$, which are natural transformations between
contravariant functors. These natural transformations satisfy the following
naturality condition:
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]
where $I, J \in \ob(\cat{C}); \Gamma, \Delta \in Psh(\cat{C}); \sigma \from
\Delta \to \Gamma$, and $f \from J \to I$. $\Gamma_f$ and $\Delta_f$ denote the
morphisms that we obtain from the morphism $f$ in $\Set$ by the fact that
$\Gamma$ and $\Delta$ are functors. For instance, if $\rho \in \Gamma(I)$, then
we have a morphism $\rho \mapsto \rho f$ written as $\Gamma_f$ to the
corresponding $\rho f \in \Gamma(J)$, (we write $f$ applied on the right to
connect to the initial morphism in $\cat{C}$) since the contravariant functor
$\Gamma$ preserves morphism mappings.

Recall that every context has a morphism from itself to the empty context.
Hence we set the empty context to be the terminal object in the category of
presheaves, the terminal context. That is a functor that maps everything to
the terminal object in $\Set$, the singleton set. Indeed, we need a unique
natural transformation from every presheaf to the terminal presheaf. This
natural transformation is a family of unique morphisms to the terminal object
in $\Set$.

@@

The second projection $\mathbf{q}$ here is the one that models the variables in
an extended context. The application of it returns the current term for the
variable. Note the similarity to the variable rule in the cartesian closed
category model of STLC, as it is also modeled by a second projection there.

@@

\begin{example}
  Let us consider a representation of a term of type $A \to A$ for some $\vdash
  \Gamma~ctxt$, $\Gamma \vdash A~type$ in the presheaf model. For a CwF $(\cat{C}, F), I, J \in ob(\cat{C}), f \from J \to I$ we give the
  element of $\mathit{Ty}(\Gamma)$ corresponding to the type $A \to A$:
  \[
    (A \to A)(I, \rho) = \{\pi~|~\pi_f(u) \in A(I, \Gamma_f \rho), u \in A(I, \Gamma_f \rho)\} = \{\lambda I f u.u\}
  \]
  Note that the required equalities hold trivially for any $g \from K \to J$,
  $((\lambda I f u.u)_f u) g = (\lambda I f u.u)_{fg} (ug)$. $\Gamma, x : A
  \vdash x : A$ corresponds to $q$ in the model. Thus $\Gamma \vdash \lambda
  x.x : A -> A$ corresponds to $\lambda q$ being defined as follows:
  \[
    ((\lambda q) (I, \rho))_f u = q(\Gamma_f \rho, u) = u
  \]
\end{example}
